---
title: "Biostat 212A Homework 3"
subtitle: "Due Feb 20, 2024 @ 11:59PM"
author: "Kathy Hoang and 506333118"
date: 02/13/2024
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
knitr:
  opts_chunk: 
    fig.align: 'center'
    # fig.width: 6
    # fig.height: 4
    message: FALSE
    cache: false
---

## ISL Exercise 5.4.2 (10pts)
We will now derive the probability that a given observation is part
of a bootstrap sample. Suppose that we obtain a bootstrap sample
from a set of n observations.

### 5.4.2.a
What is the probability that the first bootstrap observation is
not the jth observation from the original sample? Justify your
answer.

**Answer**

### 5.4.2.b
What is the probability that the second bootstrap observation
is not the jth observation from the original sample?

**Answer**

### 5.4.2.c
Argue that the probability that the jth observation is not in the
bootstrap sample is (1 − (1/n))^n.

**Answer**

### 5.4.2.d
When n = 5, what is the probability that the jth observation is
in the bootstrap sample?

**Answer**

### 5.4.2.e
When n = 100, what is the probability that the jth observation
is in the bootstrap sample?

**Answer**

### 5.4.2.f
When n = 10; 000, what is the probability that the jth observation
is in the bootstrap sample?

**Answer**

### 5.4.2.g
Create a plot that displays, for each integer value of n from 1
to 100; 000, the probability that the jth observation is in the
bootstrap sample. Comment on what you observe.

**Answer**

### 5.4.2.h
We will now investigate numerically the probability that a bootstrap
sample of size n = 100 contains the jth observation. Here
j = 4. We repeatedly create bootstrap samples, and each time
we record whether or not the fourth observation is contained in
the bootstrap sample.

Comment on the results obtained.

```{r}
store <- rep (NA , 10000)
  for (i in 1:10000) {
store [i] <- sum (sample (1:100, rep = TRUE) == 4) > 0
}
mean(store)
```
**Answer**

## ISL Exercise 5.4.9 (20pts)
We will now consider the Boston housing data set, from the ISLR2
library.

### 5.4.9.a
Based on this data set, provide an estimate for the population
mean of medv. Call this estimate $/hat{mu}$.
```{r}

```

## Least squares is MLE (10pts)
Show that in the case of linear model with Gaussian errors, maximum likelihood and least squares are the same thing, and $C_p$ and AIC are equivalent.

$$

$$

## ISL Exercise 6.6.1 (10pts)

## ISL Exercise 6.6.3 (10pts)

## ISL Exercise 6.6.4 (10pts)

## ISL Exercise 6.6.5 (10pts)

## ISL Exercise 6.6.11 (30pts)

You must follow the [typical machine learning paradigm](https://ucla-biostat-212a.github.io/2024winter/slides/06-modelselection/workflow_lasso.html) to compare _at least_ 3 methods: least squares, lasso, and ridge. Report final results as

| Method | CV RMSE | Test RMSE |
|:------:|:------:|:------:|:------:|
| LS | | | |
| Ridge | | | |
| Lasso | | | |
| ... | | | |

## Bonus question (20pts)

Consider a linear regression, fit by least squares to a set of training data $(x_1, y_1), \ldots, (x_N,  y_N)$ drawn at random from a population. Let $\hat \beta$ be the least squares estimate. Suppose we have some test data $(\tilde{x}_1, \tilde{y}_1), \ldots, (\tilde{x}_M, \tilde{y}_M)$ drawn at random from the same population as the training data. If $R_{\text{train}}(\beta) = \frac{1}{N} \sum_{i=1}^N (y_i - \beta^T x_i)^2$ and $R_{\text{test}}(\beta) = \frac{1}{M} \sum_{i=1}^M (\tilde{y}_i - \beta^T \tilde{x}_i)^2$. Show that
$$
\operatorname{E}[R_{\text{train}}(\hat{\beta})] < \operatorname{E}[R_{\text{test}}(\hat{\beta})].
$$